\boldmath Disconnected pseudo-$C_\ell$ covariances for projected large-scale structure data

### 2019-06-25 draft/paper.tex
done with text

    In this paper, we will present and generalise methods developed in the context of CMB experiments to account for the impact of survey geometry on the Gaussian part of the power spectrum covariance matrix \cite{2004MNRAS.349..603E,2005MNRAS.360.1262B,2017A&A...602A..41C}, and will study in detail the performance of these methods for large-scale structure and weak lensing datasets. We have also implemented these approximations in the public code {\tt NaMaster}\footnote{\url{https://github.com/LSSTDESC/NaMaster}} \cite{2019MNRAS.484.4127A}, making the computation of accurate Gaussian covariance matrices significantly simpler for the community.

@@

    \subsection{Quantitative comparison}\label{ssec:results.quant}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/run_sph_2b_NKA_TTTEEE_reldev_eigval_1stbin.pdf}
        \caption{Eigenvalues of the single-bin covariance matrices for all power spectra involving $\delta$ and $\gamma_E$. Results are shown for the sample covariance (solid blue), the NKA estimator (solid orange) and the spin-0 approximation (dashed green). The NKA and spin-0 estimators are able to recover the covariance eigenvalues with an accuracy $\lesssim5\%$.} \label{fig:eigv_1bin}
      \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{./figures/run_sph_2b_1stbin_chi2_TT_TE_EE.pdf}
        \caption{$\chi^2$ distributions for the cases without
          $\gamma_B$-modes. We compare the distributions obtained with the
          covariance matrix computed directly from the simulated power spectra
          (blue) and the two different analytical methods: NKA (orange) and
          spin-0 (green). In addition, the theoretical $\chi^2$ distribution
          has been included (dashed red). The distribution extracted from the
          simulation follows the theoretical expectation almost perfectly in
          all cases. In the cases of $\delta\times\gamma_E$ and
          $\gamma_E\times\gamma_E$, we observe small shifts in the peak
          $\chi^2$, of $\sim2$-$4\%$, while their width is recovered
          accurately (a difference $\lesssim 2\%$). We will show that these
          differences are due to the inaccuracy of the analytical
          approximations on the largest scales, and that they have a
          completely negligible impact on the final cosmological parameter
          constraints.} 
        \label{fig:chi2_1bin}
      \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/run_sph_2b_Spin0_NKA_TTTEEE_Full_chi2.pdf}
        \caption{Same as Figure \ref{fig:chi2_1bin} for the combination of all
          correlations between $\delta$ and $\gamma_E$ for two redshift bins,
          where similar conclusions hold. In this case, the distribution
          shift and width difference is less than a $3\%$ and $7\%$,
          respectively.}
        \label{fig:chi2_2bins}
      \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{./figures/contaminants_diag_cov.pdf}
        \caption{Diagonal of the covariance matrix estimated from simulations without contaminants (solid blue), with 200 contaminants deprojected (dashed orange) and the NKA estimator (solid green). The differences between the simulated cases are significantly smaller than those associated with the NKA estimator. The NKA method is therefore equally valid to approximate the covariance even in the presence of contaminant deprojection.}
        \label{fig:diag_conts}
      \end{figure}
      In order to quantify the validity of our analytical approximations, we need to compare the NKA covariance with the sample covariance estimated from the simulations. However, comparing two matrices is not as straightforward as comparing their elements one by one. The covariance between far-away bandpowers is expected to be very close to zero, and therefore a direct comparison of those elements would easily yield large relative differences simply due to the statistical noise in the sample covariance matrix. We will therefore quantify the differences between the different covariances making use of scalar quantities formed from them. The impact of the analytical approximations on the final parameter constraints will then be described in detail in Section \ref{ssec:results.cosmo}.

@@

      Another scalar quantity that can be used to compare different covariances is the $\chi^2$. For a given random data vector ${\bf d}$ with mean ${\bf m}$ and covariance matrix ${\rm Cov}$, this is given by
      \begin{equation}
        \chi^2=\left({\bf d}-{\bf m}\right)^T\cdot{\rm Cov}^{-1}\cdot\left({\bf d}-{\bf m}\right).
      \end{equation}
      We compute this quantity for a data vector ${\bf d}$ composed of
      different auto- and cross-correlations for each of the 20,000 Gaussian
      simulations, with ${\bf m}$ given by the mean over all simulations and
      different choices of covariance matrix. Figure \ref{fig:chi2_1bin} shows
      the distribution of $\chi^2$ values for the three non-zero power spectra
      in the case of a single bin: $\delta$-$\delta$, $\delta$-$\gamma_E$ and
      $\gamma_E$-$\gamma_E$. The histograms show the distribution for the
      sample covariance matrix (blue), and the analytical NKA and Spin-0
      estimators (orange and green, respectively). We additionally plot the theoretical $\chi^2$ distribution under the assumption that the underlying data vector is Gaussianly distributed (red dashed lines). In the simplest case of purely spin-0 quantities (leftmost panel), we find an excellent agreement between the different distributions. In the cases involving the spin-2 fields, we see noticeable differences between the distributions found with the sample covariance and the approximate ones.
      These differences are small, corresponding to less of a $2$ and $4\%$
      shift in the mean $\chi^2$ for the $\delta$-$\gamma_E$ and
      $\gamma_E$-$\gamma_E$ cases, respectively, and a negligible variation in
      the width of the distributions. We therefore expect these differents to
      have a negligible effect on the posterior parameter distributions, as we
      show explicitly in Section \ref{ssec:results.cosmo}. The fact that these
      differences appear only for power spectra involving spin-2 quantities
      indicate that the NKA and spin-0 methods are imperfect at describing the
      additional mode coupling caused in the presence of a mask for
      higher-spin fields. This is rather obvious in the case of the spin-0
      approximation but, interestingly, we find that the NKA and Spin-0
      predictions yield results that are almost indistinguishable from each
      other. We therefore conclude that the additional approximation made in
      the NKA method for spin-2 fields -- neglecting the spatial derivatives
      of the mask -- is effectively equivalent to ignoring the spin nature of
      the fields involved \textcolor{red}{for the $\delta$ and $\gamma_E$ cases. The B
        modes break this similarity and worsen the spin-0 approximation,
        yielding, in the most extreme case ($\gamma_b$- $\gamma_b$), around an
        order of magnitude worse results than the NKA method}. For
        completeness, Figure \ref{fig:chi2_2bins} shows the distribution of
        $\chi^2$ values for a data vector composed of all possible auto- and
        cross-correlations of $\delta$ and $\gamma_E$ for the case of two
        redshift bins, where similar conclusions hold.

@@

      Figure \ref{fig:diag_conts} serves also to illustrate another important
      point. In all our tests we find that the largest differences between the
      sample and NKA covariances occur on large scales. For $\delta$-$\delta$
      correlations, the effect is limited to the first few multipoles
      ($\ell\lesssim10$), while for $\gamma_E$-$\gamma_E$ we are only able to
      recover the sample covariance errors within $5\%$ for $\ell\gtrsim40$.
      This is the main source of the small mismatch observed in
      Figure~\ref{fig:chi2_1bin}. Note, however, that most of the cosmological
      information is obtained from the higher multipoles, due to their higher
      statistical weight, and, as we will show in the next section, the effect on the final parameter constraints is negligible. If accurate covariances bars are needed on these large scales for cosmic shear, they can be estimated alternatively making use of fast, low resolution simulations (e.g. {\tt HEALPix} $N_{\rm side}=64$ maps).

    \subsection{Impact on parameter estimation}\label{ssec:results.cosmo}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/contours_2bin.pdf}
        \caption{1$\sigma$ and 2$\sigma$ contours for $\Omega_M$ and $\sigma_8$ found from two different estimates of the power spectrum likelihood for a dataset containing all possible cross-power spectra between $\delta$ and $\gamma_E$ for two redshift bins. The filled contours corresponds to the parameter likelihood evaluated using the NKA estimator for the power spectrum covariance matrix, while the dashed black contours correspond to the same calculation done with the sample covariance matrix corrected by the factor in Eq. \ref{eq:hartlap}. In both cases we assume a Gaussian likelihood. We find a remarkably good agreement between both likelihoods, highlighting the negligible impact of the approximations involved in the NKA method on the final parameter estimates.} \label{fig:contours}
      \end{figure}
      Ultimately, the most important test to judge the accuracy of the analytical covariance matrix estimators implemented here is to study their impact on the posterior distribution of cosmological parameters derived from power spectrum measurements. Assuming flat priors, the posterior distribution for parameters $\vec{\theta}$ is simply given by:
      \begin{equation}\label{eq:parpos}
        -2\log p(\vec{\theta}|\hat{\bf C})=\left(\hat{\bf C}-{\bf C}(\vec{\theta})\right)^T\cdot{\rm Cov}^{-1}\cdot\left(\hat{\bf C}-{\bf C}(\vec{\theta})\right)+{\rm const.},
      \end{equation}
      where $\hat{\bf C}$ is a vector of power spectrum measurements, ${\rm Cov}$ is their covariance matrix, and ${\bf C}(\vec{\theta})$ is their theoretical prediction for parameters $\vec{\theta}$.

      We explore $\log p(\vec{\theta}|\hat{\bf C})$ for of two parameters
      $\vec{\theta}\equiv(\Omega_M,\sigma_8)$, for a data vector composed of
      all possible auto- and cross-correlations between $\delta$ and
      $\gamma_E$ in the case of two redshift bins. In this simple
      two-dimensional scenario, we simply sample the distribution in a regular grid of 100 by 100 points for each parameter. We construct a data vector $\hat{\bf C}$ from the theoretical prediction for the experimental setup described in Section \ref{ssec:results.sims}, and produce theoretical predictions for it at each grid point using {\tt CCL}. Note that, when evaluating the posterior for the sample covariance matrix, one needs to correct for the finite number of simulations used to construct the covariance. In most situations this can be done simply by rescaling the inverse covariance matrix by a factor  given by \cite{2007A&A...464..399H}
      \begin{equation}\label{eq:hartlap}
        {\rm Cov}^{-1}\,\longrightarrow \frac{N_s-2-N_{\rm data}}{N_s-1}\,{\rm Cov}^{-1},
      \end{equation}
      where $N_s=20,000$ is the number of samples used to estimate the covariance, and $N_{\rm data}=3510$ is the number of data points.

      The 68\% and 95\% confidence level contours associated with the
      posterior distributions for the sample covariance and the NKA estimator
      are shown in Figure~\ref{fig:contours}. We find that both distributions agree with each other remarkably well, and that the $1\sigma$ errors for each parameter agree for both covariances up to $0.3\%$. Note that, since we have not included any statistical noise in the data vector $\hat{\bf C}$, the relative bias between both distributions is zero by construction (since Eq. \ref{eq:parpos} is bounded from below by zero). When adding Gaussian statistical noise compatible with the sample covariance matrix, we observe small differences (smaller than 0.3$\sigma$) in the best fit parameters found with both covariance matrices. These differences, however, are not systematic, and we have verified that averaging over several noise realizations does not yield biased best-fit parameters.

@@

  \appendix
  \section{Flat sky}\label{app:flat}
    \begin{figure}
      \centering
      \includegraphics[width=0.6\textwidth]{./figures/mask-lss_flat1.pdf}
      \caption{Flat-sky mask used in the study of the analytical methods presented in this work in the flat-sky regime. The mask was constructed from the bright-object mask distributed with the first data release of the HSC collaboration \cite{2018PASJ...70S...4A} for the VVDS field. As in the full-sky case, it is used both for galaxy clustering and shear for simplicity.} \label{fig:mask_flat}
    \end{figure}

>>>>

    In this paper, we will present and generalise methods developed in the context of CMB experiments to account for the impact of survey geometry on the Gaussian part of the power spectrum covariance matrix \cite{2004MNRAS.349..603E,2005MNRAS.360.1262B,2017A&A...602A..41C}\footnote{See also \cite{2019JCAP...01..016L} for a similar application to the problem of 3D power spectrum covariances.}, and will study in detail the performance of these methods for large-scale structure and weak lensing datasets. We have also implemented these approximations in the public code {\tt NaMaster}\footnote{\url{https://github.com/LSSTDESC/NaMaster}} \cite{2019MNRAS.484.4127A}, making the computation of accurate Gaussian covariance matrices significantly simpler for the community.

@@

    \subsection{Quantitative comparison}\label{ssec:results.quant}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/run_sph_2b_NKA_TTTEEE_reldev_eigval_1stbin.pdf}
        \caption{Eigenvalues of the single-bin covariance matrices for all power spectra involving $\delta$ and $\gamma_E$. Results are shown for the sample covariance (solid blue), the NKA estimator (solid orange) and the spin-0 approximation (dashed green). The NKA and spin-0 estimators are able to recover the covariance eigenvalues with an accuracy $\lesssim5\%$.} \label{fig:eigv_1bin}
      \end{figure}
      In order to quantify the validity of our analytical approximations, we need to compare the NKA covariance with the sample covariance estimated from the simulations. However, comparing two matrices is not as straightforward as comparing their elements one by one. The covariance between far-away bandpowers is expected to be very close to zero, and therefore a direct comparison of those elements would easily yield large relative differences simply due to the statistical noise in the sample covariance matrix. We will therefore quantify the differences between the different covariances making use of scalar quantities formed from them. The impact of the analytical approximations on the final parameter constraints will then be described in detail in Section \ref{ssec:results.cosmo}.

@@

      Another scalar quantity that can be used to compare different covariances is the $\chi^2$. For a given random data vector ${\bf d}$ with mean ${\bf m}$ and covariance matrix ${\rm Cov}$, this is given by
        \chi^2=\left({\bf d}-{\bf m}\right)^T\cdot{\rm Cov}^{-1}\cdot\left({\bf d}-{\bf m}\right).
        \includegraphics[width=\textwidth]{./figures/run_sph_2b_1stbin_chi2_TT_TE_EE.pdf}
        \caption{$\chi^2$ distributions for the cases without $\gamma_B$-modes. We compare the distributions obtained with the covariance matrix computed directly from the simulated power spectra (blue) and the two different analytical methods: NKA (orange) and spin-0 (green). In addition, the theoretical $\chi^2$ distribution has been included (dashed red). The distribution extracted from the simulation follows the theoretical expectation almost perfectly in all cases. In the cases of $\delta\times\gamma_E$ and $\gamma_E\times\gamma_E$, we observe small shifts in the peak $\chi^2$, of $\sim2$-$4\%$, while their width is recovered accurately (a difference $\lesssim 2\%$). We will show that these differences are due to the inaccuracy of the analytical approximations on the largest scales, and that they have a completely negligible impact on the final cosmological parameter constraints.} 
        \label{fig:chi2_1bin}
        \includegraphics[width=0.6\textwidth]{./figures/run_sph_2b_Spin0_NKA_TTTEEE_Full_chi2.pdf}
        \caption{Same as Figure \ref{fig:chi2_1bin} for the combination of all correlations between $\delta$ and $\gamma_E$ for two redshift bins, where similar conclusions hold. In this case, the differences in the distribution means and widths are less than $3\%$ and $7\%$, respectively.}
        \label{fig:chi2_2bins}
      We compute this quantity for a data vector ${\bf d}$ composed of different auto- and cross-correlations for each of the 20,000 Gaussian simulations, with ${\bf m}$ given by the mean over all simulations and different choices of covariance matrix. Figure \ref{fig:chi2_1bin} shows the distribution of $\chi^2$ values for the three non-zero power spectra in the case of a single bin: $\delta$-$\delta$, $\delta$-$\gamma_E$ and $\gamma_E$-$\gamma_E$. The histograms show the distribution for the sample covariance matrix (blue), and the analytical NKA and Spin-0 estimators (orange and green, respectively). We additionally plot the theoretical $\chi^2$ distribution under the assumption that the underlying data vector is Gaussianly distributed (red dashed lines). In the simplest case of purely spin-0 quantities (leftmost panel), we find an excellent agreement between the different distributions. In the cases involving the spin-2 fields, we see noticeable differences between the distributions found with the sample covariance and the approximate ones. These differences are small, corresponding to less of a $2$ and $4\%$ shift in the mean $\chi^2$ for the $\delta$-$\gamma_E$ and $\gamma_E$-$\gamma_E$ cases, respectively, and a negligible variation in the width of the distributions. We therefore expect these differents to have a negligible effect on the posterior parameter distributions, as we show explicitly in Section \ref{ssec:results.cosmo}. The fact that these differences appear only for power spectra involving spin-2 quantities indicate that the NKA and spin-0 methods are imperfect at describing the additional mode coupling caused in the presence of a mask for higher-spin fields. This is rather obvious in the case of the spin-0 approximation but, interestingly, we find that the NKA and Spin-0 predictions yield results that are almost indistinguishable from each other. We therefore conclude that the additional approximation made in the NKA method for spin-2 fields -- neglecting the spatial derivatives of the mask -- is effectively equivalent to ignoring the spin nature of the fields involved. Note, however, that this is not the case for $B$-modes, where the NKA estimator outperforms the spin-0 approximation by up to one order of magnitude, even though its accuracy is very poor (as we described in the previous section). For completeness, Figure \ref{fig:chi2_2bins} shows the distribution of $\chi^2$ values for a data vector composed of all possible auto- and cross-correlations of $\delta$ and $\gamma_E$ for the case of two redshift bins, where similar conclusions hold.

@@

      \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{./figures/contaminants_diag_cov.pdf}
        \caption{Diagonal of the covariance matrix estimated from simulations without contaminants (solid blue), with 200 contaminants deprojected (dashed orange) and the NKA estimator (solid green). The differences between the simulated cases are significantly smaller than those associated with the NKA estimator. The NKA method is therefore equally valid to approximate the covariance even in the presence of contaminant deprojection.}
        \label{fig:diag_conts}
      \end{figure}
      Figure \ref{fig:diag_conts} serves also to illustrate another important point. In all our tests we find that the largest differences between the sample and NKA covariances occur on large scales. For $\delta$-$\delta$ correlations, the effect is limited to the first few multipoles ($\ell\lesssim10$), while for $\gamma_E$-$\gamma_E$ we are only able to recover the sample covariance errors within $5\%$ for $\ell\gtrsim40$. This is the main source of the small mismatch observed in Figure~\ref{fig:chi2_1bin}. Note, however, that most of the cosmological information is obtained from the higher multipoles, due to their higher statistical weight, and, as we will show in the next section, the effect on the final parameter constraints is negligible. If accurate covariances bars are needed on these large scales for cosmic shear, they can be estimated alternatively making use of fast, low resolution simulations (e.g. {\tt HEALPix} $N_{\rm side}=64$ maps).

    \subsection{Impact on parameter estimation}\label{ssec:results.cosmo}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/contours_2bin.pdf}
        \caption{1$\sigma$ and 2$\sigma$ contours for $\Omega_M$ and $\sigma_8$ found from two different estimates of the power spectrum likelihood for a dataset containing all possible cross-power spectra between $\delta$ and $\gamma_E$ for two redshift bins. The filled contours correspond to the parameter likelihood evaluated using the NKA estimator for the power spectrum covariance matrix, while the dashed black contours correspond to the same calculation done with the sample covariance matrix corrected by the factor in Eq. \ref{eq:hartlap}. In both cases we assume a Gaussian likelihood. We find a remarkably good agreement between both likelihoods, highlighting the negligible impact of the approximations involved in the NKA method on the final parameter estimates.} \label{fig:contours}
      \end{figure}
      Ultimately, the most important test to judge the accuracy of the analytical covariance matrix estimators implemented here is to study their impact on the posterior distribution of cosmological parameters derived from power spectrum measurements. Assuming flat priors, the posterior distribution for parameters $\vec{\theta}$ is simply given by:
      \begin{equation}\label{eq:parpos}
        -2\log p(\vec{\theta}|\hat{\bf C})=\left(\hat{\bf C}-{\bf C}(\vec{\theta})\right)^T\cdot{\rm Cov}^{-1}\cdot\left(\hat{\bf C}-{\bf C}(\vec{\theta})\right)+{\rm const.},
      \end{equation}
      where $\hat{\bf C}$ is a vector of power spectrum measurements, ${\rm Cov}$ is their covariance matrix, and ${\bf C}(\vec{\theta})$ is their theoretical prediction for parameters $\vec{\theta}$.

      We explore $\log p(\vec{\theta}|\hat{\bf C})$ for of two parameters $\vec{\theta}\equiv(\Omega_M,\sigma_8)$, for a data vector composed of all possible auto- and cross-correlations between $\delta$ and $\gamma_E$ in the case of two redshift bins described in Section \ref{ssec:results.sims}. In this simple two-dimensional scenario, we simply sample the distribution in a regular grid of 100 by 100 points for each parameter. We construct a data vector $\hat{\bf C}$ from the theoretical prediction for the experimental setup described in Section \ref{ssec:results.sims}, and produce theoretical predictions for it at each grid point using {\tt CCL}. Note that, when evaluating the posterior for the sample covariance matrix, one needs to correct for the finite number of simulations used to construct the covariance. In most situations this can be done simply by rescaling the inverse covariance matrix by a factor given by \cite{2007A&A...464..399H}
      \begin{equation}\label{eq:hartlap}
        {\rm Cov}^{-1}\,\longrightarrow \frac{N_s-2-N_{\rm data}}{N_s-1}\,{\rm Cov}^{-1},
      \end{equation}
      where $N_s=20,000$ is the number of samples used to estimate the covariance, and $N_{\rm data}=3510$ is the number of data points.

      The 68\% and 95\% confidence level contours associated with the posterior distributions for the sample covariance and the NKA estimator are shown in Figure~\ref{fig:contours}. We find that both distributions agree with each other remarkably well, and that the $1\sigma$ errors for each parameter agree for both covariances up to $0.3\%$. Note that, since we have not included any statistical noise in the data vector $\hat{\bf C}$, the relative difference in the means of both distributions is zero by construction (since Eq. \ref{eq:parpos} is bounded from below by zero). When adding Gaussian statistical noise compatible with the sample covariance matrix, we observe small differences (smaller than 0.3$\sigma$) in the best fit parameters found with both covariance matrices. These differences, however, are not systematic, and we have verified that averaging over several noise realizations does not yield biased best-fit parameters.

@@

  \appendix
  \section{Flat sky}\label{app:flat}
    \begin{figure}
      \centering
      \includegraphics[width=0.6\textwidth]{./figures/mask-lss_flat1.pdf}
      \caption{Flat-sky mask used in the study of the analytical methods presented in this work in the flat-sky regime. The mask was constructed from the bright-object mask distributed with the first data release of the HSC collaboration \cite{2018PASJ...70S...4A} for the VVDS field. As in the curved-sky case, it is used both for galaxy clustering and shear for simplicity.} \label{fig:mask_flat}
    \end{figure}
